When does vapor pressure deficit drive or reduce evapotranspiration?

### 2019-02-27 doc/paper/intro.tex
about to start responding to pierre's comments

The opposing effects of increased atmospheric demand and higher
stomatal stress lead to two possible perspectives for how
evapotranspiration (ET) responds to shifts in VPD. The first, a
hydrometeorological perspective, is that higher VPD increases
atmospheric demand for water from the land surface, and this drives an
increase in evapotranspiration (ET). This perspective is particularly
relevant because potential evapotranspiration (PET), which is used in
many drought indices and hydrometeorological studies
\citep[e.g.,][]{Heim_2002, Scheff_2015}, typically only quantifies
changes in atmospheric demand and fails to account for ecosystem
response \citep{Swann_2016}. In reality, plants' stomata have evolved
to optimally regulate the exchange of water and carbon, and tend to
partially close in response to increased atmospheric dryness
\citep{Farquhar_1978, Ball_1987, Leuning_1990, MEDLYN_2011}. This
leads to a plant physiology perspective, in which an increase in VPD,
particularly in well-watered soil conditions, may actually correspond
to a decrease in ET because of stomatal closure
\citep[e.g.][]{Rigden_2017}.  In other words, the question ``When does
VPD drive or reduce ET?'' can be related to whether plant regulation
or atmospheric demand dominates ET response.

@@

As a counterpoint, one may argue that increases in ET with increasing
VPD could increase the likelihood of precipitation
\citep[e.g.,][]{Findell_2011}. However, increases in ET do not always
guarantee an increased likelihood of precipitation, which depending on
environmental conditions could cause a decrease in the likelihood of
precipitation \citep[][]{Gentine_2013}. Furthermore, any increases in
precipitation are likely to be non-local, such that plants giving up
water to the atmosphere are not guaranteed to reap the benefits of
water returned from the atmosphere to the soil. Using this subjective
logic, from an ecosystem evolutionary perspective, water stored in
soil seems to be worth much more than the chance of water returned as
precipitation.

We can use intuition about plant water conservation strategy to
hypothesize about ET response to changes in VPD. Plants and ecosystems
that evolved to conserve water, such as arid shrubs or savannah,
should be more likely to reduce ET with increasing VPD, and plants
that have evolved or have been engineered to care little about water,
such as crops, will be more likely to increase ET with increasing
VPD. Atmospheric conditions must matter as well. At the ecosystem
scale, there are limits to plant water conservation strategies. As
atmospheric demand for water (VPD) increases, ecosystems should begin
to reach their water conservation limits and might not be able to
entirely limit ET flux to the atmosphere. At this stage any further
increase in VPD will most likely drive a (limited) increase in ET,
because the increase in atmospheric demand for water overwhelms the
limited plant response to conserve water.

@@

Recent results have drastically improved our understanding of VPD's
impact on physiology, especially at the leaf
level. \citet{MEDLYN_2011} developed a model for leaf-scale stomatal
conductance ($g_s$), including VPD response, by combining an optimal
photosynthesis theory \citep{Cowan_1977} with an empirical approach,
and extended use of this model to the ecosystem scale in
\citet{Medlyn_2017}. Additionally, \citet{Zhou_2014} demonstrated that
a quantity underlying water use efficiency $\left(uWUE = \frac{GPP\;
\sqrt{VPD}}{ET}\right)$ properly captures a constant relationship
between GPP, ET, and VPD over a diurnal cycle at the ecosystem
scale. uWUE is also relatively well conserved in the growing season
across space and time, within a PFT \citep{Zhou_2015}. While stomatal
conductance parameterizations and uWUE greatly simplify complex plant
physiological processes, they still capture ecosystem behavior for
vegetated surfaces \citep{Medlyn_2017, Zhou_2014}, and are novel tools
to transparently develop intuition for the behavior of complex,
multiscale ecohydrologic systems.

In this manuscript, we leverage uWUE and recent developments in
stomatal conductance parameterizations \citep{MEDLYN_2011} to derive
the theoretical one-way response of ET to VPD with other environmental
variables properly controlled for, i.e. we develop a framework for
evaluating the partial derivative of ET with respect to VPD. For the
first time, we explicitly include VPD's full effect on stomatal
conductance, including its impact on photosynthesis. We explore the
range of possible ET responses to VPD, given parameters previously
established in peer reviewed literature. Additionally, we show the
sensitivity of the ET-VPD relationship to model and framework choice,
highlighting the importance of: 1) future research on stomatal
conductance and ecosystem scaling, and 2) thoughtful selection of
photosynthesis and stomatal conductance model in more sophisticated
land surface and earth system models.

>>>>

The opposing effects of increased atmospheric demand and higher
stomatal stress lead to two possible perspectives for how
evapotranspiration (ET) responds to shifts in VPD. The first, a
hydrometeorological perspective, is that higher VPD increases
atmospheric demand for water from the land surface, and this drives an
increase in evapotranspiration (ET). However, plants' stomata have
evolved to optimally regulate the exchange of water and carbon, and
tend to partially close in response to increased atmospheric dryness
\citep{Farquhar_1978, Ball_1987, Leuning_1990, MEDLYN_2011}. This
leads to a plant physiology perspective, in which an increase in VPD
may actually correspond to a decrease in ET because of stomatal
closure \citep[e.g.][]{Rigden_2017}.  In other words, the question
``When does VPD drive or reduce ET?'' can be related to whether plant
regulation or atmospheric demand dominates ET response.

@@

We can use intuition about plant water conservation strategy to
hypothesize about ET response to changes in VPD. Plants and ecosystems
that evolved to conserve water, such as arid shrubs, should be more
likely to reduce ET with increasing VPD, and plants that have evolved
or have been engineered to care little about water, such as crops,
will be more likely to increase ET with increasing VPD. Atmospheric
conditions must matter as well. At the ecosystem scale, there are
limits to plant water conservation strategies. As atmospheric demand
for water (VPD) increases, ecosystems should begin to reach their
water conservation limits and might not be able to entirely limit ET
flux to the atmosphere. At this stage any further increase in VPD will
most likely drive a (limited) increase in ET, because the increase in
atmospheric demand for water overwhelms the limited plant response to
conserve water.

@@

Recent results have drastically improved our understanding of VPD's
impact on physiology, especially at the leaf
level. \citet{MEDLYN_2011} developed a model for leaf-scale stomatal
conductance ($g_s$), including VPD response, by combining an optimal
photosynthesis theory \citep{Cowan_1977} with an empirical approach,
and extended use of this model to the ecosystem scale in
\citet{Medlyn_2017}. Additionally, \citet{Zhou_2014} demonstrated that
a quantity underlying water use efficiency $\left(uWUE = \frac{GPP\;
\sqrt{VPD}}{ET}\right)$ properly captures a constant relationship
between GPP, ET, and VPD over a diurnal cycle at the ecosystem
scale. uWUE is also relatively well conserved in the growing season
across space and time, within a plant functional type (PFT)
\citep{Zhou_2015}. While stomatal conductance parameterizations and
uWUE greatly simplify complex plant physiological processes, they
still capture ecosystem behavior for vegetated surfaces
\citep{Medlyn_2017, Zhou_2014}, and are novel tools to transparently
develop intuition for the behavior of complex, multiscale
ecohydrologic systems.

In this manuscript, we leverage uWUE and recent developments in
stomatal conductance parameterizations \citep{MEDLYN_2011} to derive
the theoretical one-way response of ET to VPD with other environmental
variables properly controlled for, i.e. we develop a framework for
evaluating the partial derivative of ET with respect to VPD. For the
first time, we explicitly include VPD's full effect on stomatal
conductance, including its impact on photosynthesis. We explore the
range of possible ET responses to VPD, given parameters previously
established in peer reviewed literature. Additionally, we show the
sensitivity of the ET-VPD relationship to model and framework choice,
highlighting the importance of: 1) future research on stomatal
conductance and ecosystem scale modeling, and 2) thoughtful selection
of photosynthesis and stomatal conductance models in more
sophisticated land surface and earth system models.
